valr: Reproducible Genome Interval Analysis in R

### 2017-06-20 inst/paper/main.tex
fixed quotes and a few minor changes

%% Default: numerical citations
\usepackage[numbers]{natbib}

%% Uncomment this lines for superscript citations instead
% \usepackage[super]{natbib}

@@

\title{valr: Reproducible genome interval analysis in R}
\author[1]{Kent A. Riemondy}
\author[1]{Austin Gillen}
\author[2]{Ryan M. Sheridan}
\author[2]{Yinni Yu}
\author[3]{Christopher G. Bennett}
\author[1,2,*]{Jay R. Hesselberth}

@@

	A routine bioinformatic task is the analysis of the relationships between sets of genomic intervals including the identification of DNA variants within genetic regions, annotation of regions enriched for nucleic acid binding proteins, and computation of read density within a set of exons. Command-line tools for interval analysis such as BEDtools \cite{quinlan_bedtools:_2010} and BEDOPS \cite{neph_bedops:_2012} enable analyses of genome-wide datasets and are keys components of analysis pipelines. Analyses with these tools commonly combine processing intervals on the command-line with visualization and statistical analysis in R. However, the need to master both the command-line and R hinders exploratory data analysis, and the development reproducible research workflows built in the RMarkdown framework.

	Existing R packages developed for interval analysis include IRanges \cite{lawrence_software_2013}, bedr \cite{haider_bedr_2016}, and GenometriCorr \cite{favorov_exploring_2012}. IRanges is a Bioconductor package that provides  interval classes and methods to perform interval arithmetic and is used by many Bioconductor packages. bedr is a CRAN-distributed package that provides wrapper R functions to call the BEDTools, BedOPS, and tabix command-line utilities, providing out-of-memory support for interval analysis. Finally,  GenometriCorr provides a set of statistical tests to determine the relationships between interval sets using IRanges data structures. These packages provide  functionality for processing and statistical inference of interval data, but they do not easily integrate with interactive data visualization features developed for R including Shiny\cite{chang_shiny_2017}, and the RStudio IDE. Moreover, the S4 data structures implemented by IRanges do not easily interface with the recent advances provided by the popular tidyverse suite of data processing and visualization tools (e.g., dplyr, purrr, broom, and ggplot2) \cite{wickham_tidyverse_2017}. We sought to develop an flexible R package for genomic interval arithmetic built to incorporate new R programming, visualization, and interactivity features.

@@

\subsection*{Implementation}
valr is an R package that makes extensive use of dplyr, a flexible and high-performance framework for data manipulation in R \cite{wickham_dplyr_2016}. Compute intensive functions in valr are written in C++ using Rcpp to enable fluid interactive analysis of large datasets\cite{eddelbuettel_rcpp_2011}. Interval intersections and related operations use an interval tree algorithm to efficiently search for overlapping intervals\cite{cormen_2001}. BED files are imported and handled in R as data\_frame objects, requiring minimal pre or post-processing to integrate with additional R packages or command line tools.

\subsection*{Operation}
valr is distributed as part of the CRAN R package repository and is compatible with Mac OS X, Windows, and major Linux operating systems. Package dependencies and system requirements are documented in the valr CRAN repository (\url{https://cran.r-project.org/web/packages/valr/index.html}).

\section*{Use Cases}
To demonstrate the functionality and utility of \texttt{valr}, presented below is a basic tutorial for using valr and additional common use cases for genomic interval analysis.

@@

\texttt{valr} provides a set of functions to read BED, BEDgraph, and VCF formats into R as convenient \texttt{tibble} data\_frame objects. All tbls have \texttt{chrom}, \texttt{start}, and \texttt{end} columns, and tbls from multi-column formats have additional pre-determined column names. Standards methods for importing data (e.g. \texttt{read.table}, \texttt{readr::read\_tsv}) are also supported provided the constructed dataframes contain the requisite column names (\texttt{chrom}, \texttt{start}, \texttt{end}). Additionally \texttt{valr} supports connections to remote databases to programatically access the UCSC and Ensembl databases via the \texttt{db\_ucsc} and \texttt{db\_ensembl} functions.

@@

The functions in \texttt{valr} have similar names to their \texttt{BEDtools} counterparts, and so will be familiar to users of the \texttt{BEDtools} suite. Also, similar to pybedtools\cite{dale_pybedtools:_2011}, a python wrapper for \texttt{BEDTools}, \texttt{valr} has a terse syntax. For example, shown below is a demonstration of how to find all intergenic SNPs within 1 kilobase of genes using \texttt{valr}.

@@

\NormalTok{snps <-}\StringTok{ }\KeywordTok{read_bed}\NormalTok{(}\KeywordTok{valr_example}\NormalTok{(}\StringTok{'hg19.snps147.chr22.bed.gz'}\NormalTok{), }\DataTypeTok{n_fields =} \DecValTok{6}\NormalTok{)}
\NormalTok{genes <-}\StringTok{ }\KeywordTok{read_bed}\NormalTok{(}\KeywordTok{valr_example}\NormalTok{(}\StringTok{'genes.hg19.chr22.bed.gz'}\NormalTok{), }\DataTypeTok{n_fields =} \DecValTok{6}\NormalTok{)}

@@

\NormalTok{nearby %>%}
\StringTok{  }\KeywordTok{select}\NormalTok{(}\KeywordTok{starts_with}\NormalTok{(}\StringTok{'name'}\NormalTok{), .overlap, .dist) %>%}
\StringTok{  }\KeywordTok{filter}\NormalTok{(}\KeywordTok{abs}\NormalTok{(.dist) <}\StringTok{ }\DecValTok{1000}\NormalTok{)}
\CommentTok{#> # A tibble: 285 x 4}
\CommentTok{#>         name.x            name.y .overlap .dist}
\CommentTok{#>          <chr>             <chr>    <int> <int>}
\CommentTok{#>  1   rs2261631             P704P        0  -267}
\CommentTok{#>  2 rs570770556             POTEH        0  -912}
\CommentTok{#>  3 rs538163832             POTEH        0  -952}
\CommentTok{#>  4   rs9606135            TPTEP1        0  -421}
\CommentTok{#>  5  rs11912392 ANKRD62P1-PARP4P3        0   104}
\CommentTok{#>  6   rs8136454          BC038197        0   355}
\CommentTok{#>  7   rs5992556              XKR3        0  -455}
\CommentTok{#>  8 rs114101676              GAB4        0   473}
\CommentTok{#>  9  rs62236167             CECR7        0   261}
\CommentTok{#> 10   rs5747023             CECR1        0  -386}
\CommentTok{#> # ... with 275 more rows}
\end{Highlighting}

\subsubsection*{Visual documentation}\label{visual-documentation}
By conducting interval arithmetic entirely in R, \texttt{valr} is also an effective teaching tool for introducing interval analysis to early-stage analysts without requiring familiarity with both command-line tools and R. To aid in demonstrating the interval operations available in \texttt{valr}, we developed the \texttt{bed\_glyph()} tool which produces plots demonstrating the input and output of operations in \texttt{valr} in a manner similar to those found in the \texttt{BEDtools} documentation.
Shown below is the code required to produce glyphs displaying the results of intersecting \texttt{x} and \texttt{y} intervals with \texttt{bed\_intersect()} , and the result of merging \texttt{x} intervals with \texttt{bed\_merge()} (Figure 1 \ref{fig:Figure 1}).

\begin{Highlighting}[]
\NormalTok{x <-}\StringTok{ }\NormalTok{tibble::}\KeywordTok{tribble}\NormalTok{(}
  \NormalTok{~chrom, ~start, ~end,}
  \StringTok{'chr1'}\NormalTok{, }\DecValTok{25}\NormalTok{,     }\DecValTok{50}\NormalTok{,}
  \StringTok{'chr1'}\NormalTok{, }\DecValTok{100}\NormalTok{,    }\DecValTok{125}
\NormalTok{)}

\NormalTok{y <-}\StringTok{ }\NormalTok{tibble::}\KeywordTok{tribble}\NormalTok{(}
  \NormalTok{~chrom, ~start, ~end,}
  \StringTok{'chr1'}\NormalTok{, }\DecValTok{30}\NormalTok{,     }\DecValTok{75}
\NormalTok{)}

@@

\begin{Highlighting}[]
\NormalTok{x <-}\StringTok{ }\NormalTok{tibble::}\KeywordTok{tribble}\NormalTok{(}
  \NormalTok{~chrom, ~start, ~end,}
  \StringTok{'chr1'}\NormalTok{,      }\DecValTok{1}\NormalTok{,      }\DecValTok{50}\NormalTok{,}
  \StringTok{'chr1'}\NormalTok{,      }\DecValTok{10}\NormalTok{,     }\DecValTok{75}\NormalTok{,}
  \StringTok{'chr1'}\NormalTok{,      }\DecValTok{100}\NormalTok{,    }\DecValTok{120}
\NormalTok{)}

\KeywordTok{bed_glyph}\NormalTok{(}\KeywordTok{bed_merge}\NormalTok{(x))}
\end{Highlighting}
% * <jay.hesselberth@gmail.com> 2017-06-10T17:17:56.304Z:
%
% Is it possible to insert the figure inline in the document?
% * KR
% Yes, but they request that figures are listed at the end of the document.
% Once the document is formatted by F1000 I think that the document will look similar to this article:
% https://f1000research.com/articles/5-2122/v2
% ^.

@@

The \texttt{group\_by} function in dplyr can be used to perform functions
on subsets of single and multiple \texttt{data\_frame}s. Functions in
\texttt{valr} leverage grouping to enable a variety of comparisons. For
example, intervals can be grouped by \texttt{strand} to perform
comparisons among intervals on the same strand.

\begin{Highlighting}[]
\NormalTok{x <-}\StringTok{ }\NormalTok{tibble::}\KeywordTok{tribble}\NormalTok{(}
  \NormalTok{~chrom, ~start, ~end, ~strand,}
  \StringTok{'chr1'}\NormalTok{, }\DecValTok{1}\NormalTok{,      }\DecValTok{100}\NormalTok{,  }\StringTok{'+'}\NormalTok{,}
  \StringTok{'chr1'}\NormalTok{, }\DecValTok{50}\NormalTok{,     }\DecValTok{150}\NormalTok{,  }\StringTok{'+'}\NormalTok{,}
  \StringTok{'chr2'}\NormalTok{, }\DecValTok{100}\NormalTok{,    }\DecValTok{200}\NormalTok{,  }\StringTok{'-'}
\NormalTok{)}

\NormalTok{y <-}\StringTok{ }\NormalTok{tibble::}\KeywordTok{tribble}\NormalTok{(}
  \NormalTok{~chrom, ~start, ~end, ~strand,}
  \StringTok{'chr1'}\NormalTok{, }\DecValTok{50}\NormalTok{,     }\DecValTok{125}\NormalTok{,  }\StringTok{'+'}\NormalTok{,}
  \StringTok{'chr1'}\NormalTok{, }\DecValTok{50}\NormalTok{,     }\DecValTok{150}\NormalTok{,  }\StringTok{'-'}\NormalTok{,}
  \StringTok{'chr2'}\NormalTok{, }\DecValTok{50}\NormalTok{,     }\DecValTok{150}\NormalTok{,  }\StringTok{'+'}
\NormalTok{)}

@@

\subsubsection*{API}\label{api}
The major functions available in valr are shown in Table \ref{tab:tools}.

@@

\begin{Highlighting}[]
\CommentTok{# `valr_example()` identifies the path of example files in the valr package}
\NormalTok{bedfile <-}\StringTok{ }\KeywordTok{valr_example}\NormalTok{(}\StringTok{'genes.hg19.chr22.bed.gz'}\NormalTok{)}
\NormalTok{genomefile <-}\StringTok{ }\KeywordTok{valr_example}\NormalTok{(}\StringTok{'hg19.chrom.sizes.gz'}\NormalTok{)}
\NormalTok{bgfile  <-}\StringTok{ }\KeywordTok{valr_example}\NormalTok{(}\StringTok{'hela.h3k4.chip.bg.gz'}\NormalTok{)}

@@

\begin{Highlighting}[]
\CommentTok{# generate 1 bp TSS intervals, `+` strand only}
\NormalTok{tss <-}\StringTok{ }\NormalTok{genes %>%}
\StringTok{  }\KeywordTok{filter}\NormalTok{(strand ==}\StringTok{ '+'}\NormalTok{) %>%}
\StringTok{  }\KeywordTok{mutate}\NormalTok{(}\DataTypeTok{end =} \NormalTok{start +}\StringTok{ }\DecValTok{1}\NormalTok{)}

@@

Finally, these summary statistics are used to construct a plot that
illustrates histone density surrounding TSSs (Figure 2\ref{fig:Figure 2}).

@@

\KeywordTok{ggplot}\NormalTok{(res, }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =} \NormalTok{.win_id, }\DataTypeTok{y =} \NormalTok{win_mean)) +}
\StringTok{  }\KeywordTok{geom_point}\NormalTok{() +}\StringTok{ }\KeywordTok{geom_pointrange}\NormalTok{(sd_limits) +}\StringTok{ }
\StringTok{  }\KeywordTok{scale_x_continuous}\NormalTok{(}\DataTypeTok{labels =} \NormalTok{x_labels, }\DataTypeTok{breaks =} \NormalTok{x_breaks) +}\StringTok{ }
\StringTok{  }\KeywordTok{xlab}\NormalTok{(}\StringTok{'Position (bp from TSS)'}\NormalTok{) +}\StringTok{ }\KeywordTok{ylab}\NormalTok{(}\StringTok{'Signal'}\NormalTok{) +}\StringTok{ }
\StringTok{  }\KeywordTok{ggtitle}\NormalTok{(}\StringTok{'Human H3K4me3 signal near transcription start sites'}\NormalTok{) +}
\StringTok{  }\KeywordTok{theme_classic}\NormalTok{()}
\end{Highlighting}

@@

\begin{Highlighting}[]
\KeywordTok{library}\NormalTok{(tidyverse, }\DataTypeTok{warn.conflicts =} \NormalTok{F)}

\NormalTok{repeats <-}\StringTok{ }\KeywordTok{read_bed}\NormalTok{(}\KeywordTok{valr_example}\NormalTok{(}\StringTok{'hg19.rmsk.chr22.bed.gz'}\NormalTok{), }\DataTypeTok{n_fields =} \DecValTok{6}\NormalTok{) }
\NormalTok{genome <-}\StringTok{ }\KeywordTok{read_genome}\NormalTok{(}\KeywordTok{valr_example}\NormalTok{(}\StringTok{'hg19.chrom.sizes.gz'}\NormalTok{))}
\NormalTok{genes <-}\StringTok{ }\KeywordTok{read_bed12}\NormalTok{(}\KeywordTok{valr_example}\NormalTok{(}\StringTok{'hg19.refGene.chr22.bed.gz'}\NormalTok{))}
\CommentTok{# convert bed12 to bed with exons}
\NormalTok{exons <-}\StringTok{ }\KeywordTok{bed12_to_exons}\NormalTok{(genes)}

\CommentTok{# function to repeat interval shuffling}
\NormalTok{shuffle_intervals <-}\StringTok{ }\NormalTok{function(n, .data, genome) \{}
  \KeywordTok{replicate}\NormalTok{(n, }\KeywordTok{bed_shuffle}\NormalTok{(.data, genome), }\DataTypeTok{simplify =} \OtherTok{FALSE}\NormalTok{) %>%}
\StringTok{    }\KeywordTok{bind_rows}\NormalTok{(}\DataTypeTok{.id =} \StringTok{'rep'}\NormalTok{) %>%}
\StringTok{    }\KeywordTok{group_by}\NormalTok{(rep) %>%}\StringTok{ }\KeywordTok{nest}\NormalTok{()}
\NormalTok{\}}
\NormalTok{nreps <-}\StringTok{ }\DecValTok{100}
\NormalTok{shuffled <-}\StringTok{ }\KeywordTok{shuffle_intervals}\NormalTok{(}\DataTypeTok{n =} \NormalTok{nreps, repeats, genome) %>%}
\StringTok{  }\KeywordTok{mutate}\NormalTok{(}\DataTypeTok{jaccard =} \NormalTok{data %>%}
\StringTok{           }\KeywordTok{map}\NormalTok{(bed_jaccard, repeats) %>%}
\StringTok{           }\KeywordTok{map_dbl}\NormalTok{(}\StringTok{"jaccard"}\NormalTok{))}
\NormalTok{shuffled}
\CommentTok{#> # A tibble: 100 x 3}
\CommentTok{#>      rep                  data      jaccard}
\CommentTok{#>    <chr>                <list>        <dbl>}
\CommentTok{#>  1     1 <tibble [10,000 x 6]> 0.0004337515}
\CommentTok{#>  2     2 <tibble [10,000 x 6]> 0.0003501136}
\CommentTok{#>  3     3 <tibble [10,000 x 6]> 0.0006080318}
\CommentTok{#>  4     4 <tibble [10,000 x 6]> 0.0005768465}
\CommentTok{#>  5     5 <tibble [10,000 x 6]> 0.0002537739}
\CommentTok{#>  6     6 <tibble [10,000 x 6]> 0.0004480599}
\CommentTok{#>  7     7 <tibble [10,000 x 6]> 0.0001777942}
\CommentTok{#>  8     8 <tibble [10,000 x 6]> 0.0003448734}
\CommentTok{#>  9     9 <tibble [10,000 x 6]> 0.0003303327}
\CommentTok{#> 10    10 <tibble [10,000 x 6]> 0.0003312376}
\CommentTok{#> # ... with 90 more rows}

@@

\subsection*{Benchmarking against bedtools}
In order to ensure that valr performs fast enough to enable interactive analysis, key functionality is implemented in C++. To test the speed of major valr functions we generated two data\_frames containing 1 million random 1 kilobase intervals derived from the human genome. Most of the major valr functions complete execution in less than 1 second, demonstrating that valr can process large interval datasets efficiently (Figure 3A\ref{fig:Figure 3}).

>>>>

%% Default: numerical citations
%\usepackage[numbers]{natbib}

%% Uncomment this lines for superscript citations instead
\usepackage[super]{natbib}

@@

\title{valr: Reproducible Genome Interval Analysis in R}
\author[1]{Kent A. Riemondy}
\author[2]{Ryan M. Sheridan}
\author[1]{Austin Gillen}
\author[2]{Yinni Yu}
\author[3]{Christopher G. Bennett}
\author[1,2,*]{Jay R. Hesselberth}

@@

	A routine bioinformatic task is the analysis of the relationships between sets of genomic intervals including the identification of DNA variants within protein coding regions, annotation of regions enriched for nucleic acid binding proteins, and computation of read density within a set of exons. Command-line tools for interval analysis such as \texttt{BEDtools} \cite{quinlan_bedtools:_2010} and \texttt{BEDOPS} \cite{neph_bedops:_2012} enable analyses of genome-wide datasets and are key components of analysis pipelines. Analyses with these tools commonly combine processing intervals on the command-line with visualization and statistical analysis in R. However, the need to master both the command-line and R hinders exploratory data analysis, and the development of reproducible research workflows built in the RMarkdown framework.

	Existing R packages developed for interval analysis include \texttt{IRanges} \cite{lawrence_software_2013}, \texttt{bedr} \cite{haider_bedr_2016}, and \texttt{GenometriCorr} \cite{favorov_exploring_2012}. \texttt{IRanges} is a Bioconductor package that provides interval classes and methods to perform interval arithmetic and is used by many Bioconductor packages. \texttt{bedr} is a CRAN-distributed package that provides wrapper R functions to call the \texttt{BEDtools}, \texttt{BEDOPS}, and \texttt{tabix} command-line utilities, providing out-of-memory support for interval analysis. Finally, \texttt{ GenometriCorr} provides a set of statistical tests to determine the relationships between interval sets using \texttt{IRanges} data structures. These packages provide functionality for processing and statistical inference of interval data, however they require a detailed understanding of S4 classes (\texttt{IRanges}) or the installation of external command-line dependencies (\texttt{bedr}). Additionally, these packages do not easily integrate with the recent advances provided by the popular \texttt{tidyverse} suite of data processing and visualization tools (e.g. \texttt{dplyr}, \texttt{purrr}, \texttt{broom} and \texttt{ggplot2}) \cite{wickham_tidyverse_2017}. We therefore sought to develop a flexible R package for genomic interval arithmetic built to incorporate new R programming, visualization, and interactivity features.

@@

\subsection*{Implementation}
\texttt{valr} is an R package that makes extensive use of \texttt{dplyr}, a flexible and high-performance framework for data manipulation in R \cite{wickham_dplyr_2016}. Additionally, compute intensive functions in \texttt{valr} are written in C++ using \texttt{Rcpp} to enable fluid interactive analysis of large datasets \cite{eddelbuettel_rcpp_2011}. Interval intersections and related operations use an interval tree algorithm to efficiently search for overlapping intervals \cite{cormen_2001}. BED files are imported and handled in R as \texttt{data\_frame} objects, requiring minimal pre or post-processing to integrate with additional R packages or command line tools.

\subsection*{Operation}
\texttt{valr} is distributed as part of the CRAN R package repository and is compatible with Mac OS X, Windows, and major Linux operating systems. Package dependencies and system requirements are documented in the \href{https://cran.r-project.org/web/packages/valr/index.html}{\texttt{valr} CRAN repository}.

\section*{Use Cases}
To demonstrate the functionality and utility of \texttt{valr}, presented below is a basic tutorial for using \texttt{valr} and additional common use cases for genomic interval analysis.

@@

\texttt{valr} provides a set of functions to read BED, BEDgraph, and VCF formats into R as convenient \texttt{tibble} (tbl) \texttt{data\_frame} objects. All tbls have \texttt{chrom}, \texttt{start}, and \texttt{end} columns, and tbls from multi-column formats have additional pre-determined column names. Standards methods for importing data (e.g. \texttt{read.table}, \texttt{readr::read\_tsv}) are also supported provided the constructed dataframes contain the requisite column names (\texttt{chrom}, \texttt{start}, \texttt{end}). Additionally, \texttt{valr} supports connections to remote databases to access the UCSC and Ensembl databases via the \texttt{db\_ucsc} and \texttt{db\_ensembl} functions.

@@

The functions in \texttt{valr} have similar names to their \texttt{BEDtools} counterparts, and so will be familiar to users of the \texttt{BEDtools} suite. Also, similar to pybedtools\cite{dale_pybedtools:_2011}, a python wrapper for \texttt{BEDtools}, \texttt{valr} has a terse syntax. For example, shown below is a demonstration of how to find all intergenic SNPs within 1 kilobase of genes using \texttt{valr}.

@@

\NormalTok{snps <-}\StringTok{ }\KeywordTok{read_bed}\NormalTok{(}\KeywordTok{valr_example}\NormalTok{(}\StringTok{"hg19.snps147.chr22.bed.gz"}\NormalTok{), }\DataTypeTok{n_fields =} \DecValTok{6}\NormalTok{)}
\NormalTok{genes <-}\StringTok{ }\KeywordTok{read_bed}\NormalTok{(}\KeywordTok{valr_example}\NormalTok{(}\StringTok{"genes.hg19.chr22.bed.gz"}\NormalTok{), }\DataTypeTok{n_fields =} \DecValTok{6}\NormalTok{)}

@@

\NormalTok{nearby %>%}
\StringTok{  }\KeywordTok{select}\NormalTok{(}\KeywordTok{starts_with}\NormalTok{(}\StringTok{"name"}\NormalTok{), .overlap, .dist) %>%}
\StringTok{  }\KeywordTok{filter}\NormalTok{(}\KeywordTok{abs}\NormalTok{(.dist) <}\StringTok{ }\DecValTok{1000}\NormalTok{)}
\CommentTok{#> # A tibble: 285 x 4}
\CommentTok{#>         name.x            name.y .overlap .dist}
\CommentTok{#>          <chr>             <chr>    <int> <int>}
\CommentTok{#>  1   rs2261631             P704P        0  -267}
\CommentTok{#>  2 rs570770556             POTEH        0  -912}
\CommentTok{#>  3 rs538163832             POTEH        0  -952}
\CommentTok{#>  4   rs9606135            TPTEP1        0  -421}
\CommentTok{#>  5  rs11912392 ANKRD62P1-PARP4P3        0   104}
\CommentTok{#>  6   rs8136454          BC038197        0   355}
\CommentTok{#>  7   rs5992556              XKR3        0  -455}
\CommentTok{#>  8 rs114101676              GAB4        0   473}
\CommentTok{#>  9  rs62236167             CECR7        0   261}
\CommentTok{#> 10   rs5747023             CECR1        0  -386}
\CommentTok{#> # ... with 275 more rows}
\end{Highlighting}

\subsubsection*{Visual documentation}\label{visual-documentation}
By conducting interval arithmetic entirely in R, \texttt{valr} is also an effective teaching tool for introducing interval analysis to early-stage analysts without requiring familiarity with both command-line tools and R. To aid in demonstrating the interval operations available in \texttt{valr}, we developed the \texttt{bed\_glyph()} tool which produces plots demonstrating the input and output of operations in \texttt{valr} in a manner similar to those found in the \texttt{BEDtools} documentation.
Shown below is the code required to produce glyphs displaying the results of intersecting \texttt{x} and \texttt{y} intervals with \texttt{bed\_intersect()} , and the result of merging \texttt{x} intervals with \texttt{bed\_merge()} (Figure \ref{fig:Figure 1}).

\begin{Highlighting}[]
\NormalTok{x <-}\StringTok{ }\NormalTok{tibble::}\KeywordTok{tribble}\NormalTok{(}
  \NormalTok{~chrom, ~start, ~end,}
  \StringTok{"chr1"}\NormalTok{, }\DecValTok{25}\NormalTok{,     }\DecValTok{50}\NormalTok{,}
  \StringTok{"chr1"}\NormalTok{, }\DecValTok{100}\NormalTok{,    }\DecValTok{125}
\NormalTok{)}

\NormalTok{y <-}\StringTok{ }\NormalTok{tibble::}\KeywordTok{tribble}\NormalTok{(}
  \NormalTok{~chrom, ~start, ~end,}
  \StringTok{"chr1"}\NormalTok{, }\DecValTok{30}\NormalTok{,     }\DecValTok{75}
\NormalTok{)}

@@

\begin{Highlighting}[]
\NormalTok{x <-}\StringTok{ }\NormalTok{tibble::}\KeywordTok{tribble}\NormalTok{(}
  \NormalTok{~chrom, ~start, ~end,}
  \StringTok{"chr1"}\NormalTok{,      }\DecValTok{1}\NormalTok{,      }\DecValTok{50}\NormalTok{,}
  \StringTok{"chr1"}\NormalTok{,      }\DecValTok{10}\NormalTok{,     }\DecValTok{75}\NormalTok{,}
  \StringTok{"chr1"}\NormalTok{,      }\DecValTok{100}\NormalTok{,    }\DecValTok{120}
\NormalTok{)}

\KeywordTok{bed_glyph}\NormalTok{(}\KeywordTok{bed_merge}\NormalTok{(x))}
\end{Highlighting}

@@

The \texttt{group\_by} function in \texttt{dplyr} can be used to execute functions
on subsets of single and multiple \texttt{data\_frame}s. Functions in
\texttt{valr} leverage grouping to enable a variety of comparisons. For
example, intervals can be grouped by \texttt{strand} to perform
comparisons among intervals on the same strand.

\begin{Highlighting}[]
\NormalTok{x <-}\StringTok{ }\NormalTok{tibble::}\KeywordTok{tribble}\NormalTok{(}
  \NormalTok{~chrom, ~start, ~end, ~strand,}
  \StringTok{"chr1"}\NormalTok{, }\DecValTok{1}\NormalTok{,      }\DecValTok{100}\NormalTok{,  }\StringTok{"+"}\NormalTok{,}
  \StringTok{"chr1"}\NormalTok{, }\DecValTok{50}\NormalTok{,     }\DecValTok{150}\NormalTok{,  }\StringTok{"+"}\NormalTok{,}
  \StringTok{"chr2"}\NormalTok{, }\DecValTok{100}\NormalTok{,    }\DecValTok{200}\NormalTok{,  }\StringTok{"-"}
\NormalTok{)}

\NormalTok{y <-}\StringTok{ }\NormalTok{tibble::}\KeywordTok{tribble}\NormalTok{(}
  \NormalTok{~chrom, ~start, ~end, ~strand,}
  \StringTok{"chr1"}\NormalTok{, }\DecValTok{50}\NormalTok{,     }\DecValTok{125}\NormalTok{,  }\StringTok{"+"}\NormalTok{,}
  \StringTok{"chr1"}\NormalTok{, }\DecValTok{50}\NormalTok{,     }\DecValTok{150}\NormalTok{,  }\StringTok{"-"}\NormalTok{,}
  \StringTok{"chr2"}\NormalTok{, }\DecValTok{50}\NormalTok{,     }\DecValTok{150}\NormalTok{,  }\StringTok{"+"}
\NormalTok{)}

@@

\subsubsection*{API}\label{api}
The major functions available in \texttt{valr} are shown in Table \ref{tab:tools}.

@@

\begin{Highlighting}[]
\NormalTok{bedfile <-}\StringTok{ }\KeywordTok{valr_example}\NormalTok{(}\StringTok{"genes.hg19.chr22.bed.gz"}\NormalTok{)}
\NormalTok{genomefile <-}\StringTok{ }\KeywordTok{valr_example}\NormalTok{(}\StringTok{"hg19.chrom.sizes.gz"}\NormalTok{)}
\NormalTok{bgfile  <-}\StringTok{ }\KeywordTok{valr_example}\NormalTok{(}\StringTok{"hela.h3k4.chip.bg.gz"}\NormalTok{)}

@@

\begin{Highlighting}[]
\CommentTok{# generate 1 bp TSS intervals, "+" strand only}
\NormalTok{tss <-}\StringTok{ }\NormalTok{genes %>%}
\StringTok{  }\KeywordTok{filter}\NormalTok{(strand ==}\StringTok{ "+"}\NormalTok{) %>%}
\StringTok{  }\KeywordTok{mutate}\NormalTok{(}\DataTypeTok{end =} \NormalTok{start +}\StringTok{ }\DecValTok{1}\NormalTok{)}

@@

Finally, these summary statistics are used to construct a plot that
illustrates histone density surrounding TSSs (Figure \ref{fig:Figure 2}).

@@

\NormalTok{p <-}\StringTok{ }\KeywordTok{ggplot}\NormalTok{(res, }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =} \NormalTok{.win_id, }\DataTypeTok{y =} \NormalTok{win_mean)) +}
\StringTok{  }\KeywordTok{geom_point}\NormalTok{(}\DataTypeTok{size =} \FloatTok{0.25}\NormalTok{) +}\StringTok{ }\KeywordTok{geom_pointrange}\NormalTok{(sd_limits, }\DataTypeTok{size =} \FloatTok{0.1}\NormalTok{) +}\StringTok{ }
\StringTok{  }\KeywordTok{scale_x_continuous}\NormalTok{(}\DataTypeTok{labels =} \NormalTok{x_labels, }\DataTypeTok{breaks =} \NormalTok{x_breaks) +}\StringTok{ }
\StringTok{  }\KeywordTok{xlab}\NormalTok{(}\StringTok{"Position (bp from TSS)"}\NormalTok{) +}\StringTok{ }\KeywordTok{ylab}\NormalTok{(}\StringTok{"Signal"}\NormalTok{) +}\StringTok{ }
\StringTok{  }\KeywordTok{theme_classic}\NormalTok{()}
\end{Highlighting}

@@

\begin{Highlighting}[]
\KeywordTok{library}\NormalTok{(tidyverse)}

\NormalTok{repeats <-}\StringTok{ }\KeywordTok{read_bed}\NormalTok{(}\KeywordTok{valr_example}\NormalTok{(}\StringTok{"hg19.rmsk.chr22.bed.gz"}\NormalTok{), }\DataTypeTok{n_fields =} \DecValTok{6}\NormalTok{) }
\NormalTok{genome <-}\StringTok{ }\KeywordTok{read_genome}\NormalTok{(}\KeywordTok{valr_example}\NormalTok{(}\StringTok{"hg19.chrom.sizes.gz"}\NormalTok{))}
\NormalTok{genes <-}\StringTok{ }\KeywordTok{read_bed12}\NormalTok{(}\KeywordTok{valr_example}\NormalTok{(}\StringTok{"hg19.refGene.chr22.bed.gz"}\NormalTok{))}
\CommentTok{# convert bed12 to bed with exons}
\NormalTok{exons <-}\StringTok{ }\KeywordTok{bed12_to_exons}\NormalTok{(genes)}

\CommentTok{# function to repeat interval shuffling}
\NormalTok{shuffle_intervals <-}\StringTok{ }\NormalTok{function(n, .data, genome) \{}
  \KeywordTok{replicate}\NormalTok{(n, }\KeywordTok{bed_shuffle}\NormalTok{(.data, genome), }\DataTypeTok{simplify =} \OtherTok{FALSE}\NormalTok{) %>%}
\StringTok{    }\KeywordTok{bind_rows}\NormalTok{(}\DataTypeTok{.id =} \StringTok{"rep"}\NormalTok{) %>%}
\StringTok{    }\KeywordTok{group_by}\NormalTok{(rep) %>%}\StringTok{ }\KeywordTok{nest}\NormalTok{()}
\NormalTok{\}}

\NormalTok{nreps <-}\StringTok{ }\DecValTok{100}

\NormalTok{shuffled <-}\StringTok{ }\KeywordTok{shuffle_intervals}\NormalTok{(}\DataTypeTok{n =} \NormalTok{nreps, repeats, genome) %>%}
\StringTok{  }\KeywordTok{mutate}\NormalTok{(}\DataTypeTok{jaccard =} \NormalTok{data %>%}
\StringTok{           }\KeywordTok{map}\NormalTok{(bed_jaccard, repeats) %>%}
\StringTok{           }\KeywordTok{map_dbl}\NormalTok{(}\StringTok{"jaccard"}\NormalTok{))}
\NormalTok{shuffled}
\CommentTok{#> # A tibble: 100 x 3}
\CommentTok{#>      rep                  data      jaccard}
\CommentTok{#>    <chr>                <list>        <dbl>}
\CommentTok{#>  1     1 <tibble [10,000 x 6]> 0.0003388967}
\CommentTok{#>  2     2 <tibble [10,000 x 6]> 0.0004965988}
\CommentTok{#>  3     3 <tibble [10,000 x 6]> 0.0002974843}
\CommentTok{#>  4     4 <tibble [10,000 x 6]> 0.0006899870}
\CommentTok{#>  5     5 <tibble [10,000 x 6]> 0.0004678412}
\CommentTok{#>  6     6 <tibble [10,000 x 6]> 0.0001726937}
\CommentTok{#>  7     7 <tibble [10,000 x 6]> 0.0004694941}
\CommentTok{#>  8     8 <tibble [10,000 x 6]> 0.0004660410}
\CommentTok{#>  9     9 <tibble [10,000 x 6]> 0.0006846643}
\CommentTok{#> 10    10 <tibble [10,000 x 6]> 0.0002143829}
\CommentTok{#> # ... with 90 more rows}

@@

\subsection*{Benchmarking against bedtools}
In order to ensure that \texttt{valr} performs fast enough to enable interactive analysis, key functionality is implemented in C++. To test the speed of major \texttt{valr} functions we generated two \texttt{data\_frames} containing 1 million randomly selected 1 kilobase intervals derived from the human genome. Most of the major \texttt{valr} functions complete execution in less than 1 second, demonstrating that \texttt{valr} can process large interval datasets efficiently (Figure \ref{fig:Figure 3}{A}).

We also benchmarked major \texttt{valr} functions against corresponding commands in \texttt{BEDtools}. \texttt{valr} operates on \texttt{data\_frames} already loaded into RAM, whereas \texttt{BEDtools} performs file-reading, processing, and writing. To compare \texttt{valr} against \texttt{BEDtools} we generated two BED files containing 1 million randomly selected 1 kilobase intervals derived from the human genome. For \texttt{valr} functions, we timed reading the table into R (e.g. with \texttt{read\_bed()}) and performing the respective function. For \texttt{BEDtools} commands we timed executing the command with the output written to \texttt{/dev/null}. \texttt{valr} functions performed similarly or faster than \texttt{BEDtools} commands with the exception of \texttt{bed\_map} and \texttt{bed\_fisher} (Figure \ref{fig:Figure 3}{B}).
